Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-06-17 draft.tex
continued rewrite of application section

%% indicate IEEE Member or Student Member in form indicated below
\author{P. Klemm$^{1}$, K. Lawonn$^{1}$, M. Rak$^{1}$, B. Preim$^{1}$, K. Toennies$^{1}$, K. Hegenscheid$^{2}$, H. Völzke$^{2}$, S. Oeltze$^{1}$}
\authorfooter{
%% insert punctuation at end of each item
\item
 $^1$Otto-von-Guericke University Magdeburg, Germany.
\item
 $^2$Ernst-Moritz-Arndt-University Greifswald, Germany.
}

@@

Morphometric features are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis Workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 \caption{Detailed \emph{IVA Loop} as extension from Figure~\ref{fig:WorkflowComparison}.
 %
 Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
 %
 The data are visualized in the range and domain space, which can then be brushed, yielding new groups, to be investigated using further features.
 %
 Note that adjacent steps are directly connected via feedback loops allowing for an iterative refinement and giving as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
%-------------------------------- Figure --------------------------------
\begin{figure*}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 % \includegraphics[width=3.5in]{figures/visualization}
 %\includegraphics[width=1\textwidth]{figures/visualization}
 \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
 \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization can be chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination. The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop. Dropping the \emph{gender} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
 }
 \label{fig:visualization}
\end{figure*}
%-------------------------------- /Figure -------------------------------
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, visualization and brushing.
%
A hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection from a shape-based clustering, which creates shape groups.
%
\emph{Hypothesis generation} with focus on image data starts with a shape-based clustering.
%
The feature is mapped using an automatically chosen visualization appropriate for its data type (described in detail in the following section).
%
The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
%
In our system, the visualization can either be brushed or new features can be added to the analysis. 
%
Brushing methods are subdivided using the previously described \emph{IVA} patterns.
%
Brushed regions are treated like features as they divide the subject space just like categorical features.
%
Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight features associated with selected features.
%
A sample workflow using interaction and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.

@@

The front-end is created using \texttt{Bootstrap}\footnote{Developed by Twitter, \url{http://getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%
Information visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} library \cite{D3}, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
%
The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by Nicolas Kruchten, \url{http://nicolas.kruchten.com/pivottable}}.
%
\texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} allows GPU-accelerated data rendering using \texttt{WebGL}.
%
Communication between client and server is enabled via the WebSockets protocol.
%
Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{http://mathworks.com}}, we had to access them using the \texttt{NodeJS} server.
%
We accomplished this by converting it to a parameterized standalone console application that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard out and returns it in a proper format to the client.
%
All parameter-steered console applications can be incorporated in this context.
\\\\
\section{Application}
\add{This Chapter needs to be reworked completely, since it was subject of the reviewer criticism.}
This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
%
We conducted a case study with two domain experts on a data set compiled to analyze lower back pain. 
%
This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
%
Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
%
In comparable studies, only a few shape-related features are included \cite{Lang2011}.
%
To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
%
Determining risk factors in this area can lead to \cite{Fletcher2012}:
\begin{itemize}
	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
	\item prognostic features for diagnosis and treatment of lumbar back pain, and
	\item determination of particularly affected risk groups.
\end{itemize}
%
Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
%
\subsection{The Lumbar Spine Data Set}
We divide the data set into image and non-image data.
%
There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
%
The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
%
%The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.

@@

We conducted a case study with two epidemiological domain experts.
%
Both of them are practicing clinicians with focus on epidemiological research.
%
The first clinician is an specialist in internal medicine and also the leader of the SHIP.
%
One clinician is an radiologist and responsible for the image acquisition of the SHIP and has a deep understanding on the available image data.
%
\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule into account, the case study was done completely web based.
%
Since one major reason for the focus on web technologies, no software needed to be exchanged.
%
A website link was send to the users, which they could open with their browser of choice and start using the tool.
%
To make the user input comprehensible, we observed their monitor via screen sharing.
%
Communication was enabled via webcam supported voice over ip.
%
The total setup time including installing the screen sharing application was about five minutes.
%
Video-recordings of the sessions made it possible to evaluate the sessions later on.
\com{TODO: Name der Screensharing Software mit angeben (Teamviewer)?}
\paragraph{Procedure.}
At first, controlled mouse and keyboard of the clinicians pc and demonstrated the basic functionalities of the system.
%
This included the Heatmap, the correlation view, how to introduce parameter visualization, how the color coding of differences works and the pivot tables.
%
As they understood the concepts, we handed over the mouse and keyboard control and only observed from this point on.
%
The epidemiologists were given two tasks: one hypothesis-free analysis of the data and one starting with an assumption.

% \subsection{Exploratory Analysis of the Lumbar Spine Data Set}
\subsection{Hypothesis-free analysis}
Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
%
With our tool, there are two ways to achieve this.
%
Performing a \emph{local investigation} by clustering all subjects or use the contingency matrix for a \emph{multivariate analysis}.
%
The latter was chosen by both domain experts.
%
They were excited by the possibility to look at all variables in the context of each other.
%
They stated that it "illumunates the black box of the data", making it possible to look at the data in an unbiased way.
%
Due to the dense mapping of information in the contingency matrix it was suggested to make this visualization full screen.
%
The radiologist was looking for correlations with shape related features in the data, finding that they correlate strong with each other as well as with leg pain, age, body height and hormone replacement therapy.
%

After this initial overview, both experts performed a \emph{multivariate analysis} by introducing features, such as age, waist circumference, weight or lumbar spine centerline curvature as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
%

@@

\subsection{Hypothesis-based analysis}
The radiologist performed a \emph{local investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
%
Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
%
The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
%

A second analysis was carried out on body weight.
%\com{Video 00.29. Analysis of Gender and Age - older subjects are closer to females, younger to males. This is due to shrinking of the spine in the aging process.}
\com{00.46 Second: Analysis of Female subjects via clustering, selction of cluster 2, analysis using Body weight and Age. Clustering Result if 1 gives large distance between other subjects. 00.49 How are can the differences be explained?}
\com{00.46 Feedback Body-Size/Body weight. Some features are sparse, suggestion was made to use quantile/quintile, resulting in equally occupied results. This also depends on what the user is intersted in - for clinical use the outliers may be interesting (57.30). (58.30) From perspective of a radiologists, quartile and quintiles make sense when establishing reference values for diseases; clinical relevant parameter, einflussfaktoren oder risikofaktoren zu identifizieren, je nach dem welche hypothese man verwendet.}
\com{Backpain three month > age. There are more subjects reporting backpain than those who dont. Since influences heavily the average shape - since most people have back pain, the average shape already is associated with back pain, which makes it difficult to extract parameters which reliably describe back pain. (1.05) The radiologist pointed out that for the mosaic view it would be beneficial to abstract the shape and curvature and categorize it into shapes s, i form. It was also pointed out that in the plot back pain>age, back pain can not be differenciated because the difference is mainly encoded via the age, more variables need to be added (gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview).}
\subsection{Lessons Learned}
\com{ToDo: Noch einmal auf die VDAR Results eingehen}
The radiologist pointed out the unusual strong similarity of the L3 vertebrae throughout the population.
The medical explanation is that it represents a angular point of curvature of the lumbar spine.
A seconds explanation is the use of the L3 vertebra is initialization point of the lumbar spine model \com{(00.54)}.
The domain experts see another great potential in communicating insights efficiently using the presented visualizations \com{(1.08)}.
The radiologist pointed out that the image data representations would be better comparable if they would be abstracted into representative land marks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the adjacency matrix for getting insight into a large amount of features simultaneously.
It is also useful to verify established but still controversial risk factors, such as the metabolic syndrome for coronary heart disease and whether the data set provides even more suitable risk factors.
Creating adjacency matrices for sub groups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.
%She also pointed out the benefits of an automatic categorization of the subjects into shape groups.%$$, which allows for even faster comparison.
\com{Employ a more abstract visualization of the spine to capture reliable information and enable better comparison between subjects}
%Since many large population studies incorporate medical image data this gains importance.
%\com{1.08 It was pointed out that this technique is also good for epidemiological publication because it is often more insigthfull to show these images instead of long tables. }
\com{ToDo Fazit über Web-Austausch, das Experten begeistert davon waren und dass es hervorragend funktioniert. Kleine Änderungen können sogar live programmiert werden ohne das etwas neu kompiliert werden muss (hier wäre ein Beispiel nett, z.B. Größe der Schrift irgendwo. Außerdem bringt der Browser viele Funktionen mit, wie z.B. die Suchfunktionen.)}

\com{Small User Interface Suggestions: Search-Bar for Subjects, making the elements of the contingency matrix selectable; See which Subject is currently selected}
\com{Enhancement contingency matrix: Display exact value of correlations, also information about distribution. More focus on this}
\com{Außerdem ein Anwendungsgebiet für die Matrix: Einschätzen metabolisches syndrom, eine Menge von Risikofaktoren die für Koronare Herzkrankheit Marker darstellen und die bereits länger in der Kritik sind. Hier kann getestet werden, ob es hierfür nicht bessere Indikatoren noch geben kann! (39.32)}
%\com{Also mosaic plots can be incorporated for several sub groups to see where the main differences are (for different age groups to characterize age)}
\com{1.20 More exact segmentations are useful since curvature changes are often the last change happening in a pathological process, after local deformations of the vertebrae}

% ---------------------------------------------------------------------------------------------------------------------------------------
% ---------------------------------------------------------------------------------------------------------------------------------------
% ---------------------------------------------------------------------------------------------------------------------------------------
\section{Old Application Section}
%
This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
%
We applied the presented set of techniques to a data set compiled to analyze lower back pain. 
%
This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
%
Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
%
In comparable studies, only a few shape-related features are included \cite{Lang2011}.
%
To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
%
Determining risk factors in this area can lead to \cite{Fletcher2012}:
\begin{itemize}
	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
	\item prognostic features for diagnosis and treatment of lumbar back pain, and
	\item determination of particularly affected risk groups.
\end{itemize}
% Determining risk factors in this area can lead to \cite{Fletcher2012} (evtl. weglassen, Dopplung bei Epidemiological Background?):
% \begin{itemize}
% 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
% 	\item prognostic features for diagnosis and treatment of lumbar back pain
% 	\item determination of particularly effected risk groups
% \end{itemize}
%
Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.

Data confidentiality and ethical reasons prohibit us from accessing the complete \texttt{SHIP} feature space.
%
Our epidemiological collaborators compiled a feature list, which is a tradeoff between complexity and limitations of the responsible ethics committee.
%
\subsection{The Lumbar Spine Data Set}
We divide the data set into image and non-image data.
%
There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
%
The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
%
%The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.

\subsection{Data Preprocessing} \label{application:Data Preprocessing}
The data are preprocessed as described in Section~\ref{Data Preprocessing}.
%
\paragraph{Non-Image Data.}

\subsection{Shape Visualization and Clustering}
%
The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
%
This allows for calculation of shape distance and similarity.
%
This information is used to calculate mean shapes as described in Section~\ref{Interaction- and Visualization Techniques}.

Shape distance is mapped onto color.
%
For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
\\\\
Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
%
Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
%
The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.

\subsection{Exploratory Analysis of the Lumbar Spine Data Set}
%
Expert-guided analysis assessed the suitability of our approach for supporting both hypothesis-free analysis as well as hypothesis generation. 
%
The use of \emph{IVA} patterns as presented in Section~\ref{Image Centric Cohort Study Data in Interactive Visual Analysis Context} is denoted accordingly.
%
\subsubsection{Hypothesis-Free Analysis} \label{Hypothesis free Analysis}
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/hypothesisfree}
 \caption{(a) Clustering result of all subjects. The bar chart height indicates the number of subjects in the cluster. The difference to the mean shape is color-coded, whereas blue represents no difference and red a large difference. (b) A pivot table relates cluster results to gender and employment status. Cluster 8 is striking because of the high proportion of women, while cluster 2, 3, 4 contain almost only men.}
 \label{fig:hypothesisfree}
\end{figure}
%
Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
%
Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
%
The results for this step can be seen in Figure~\ref{fig:hypothesisfree} (a).

>>>>

%% indicate IEEE Member or Student Member in form indicated below
\author{P. Klemm$^{1}$, S. Oeltze$^{1}$, K. Lawonn$^{1}$, K. Hegenscheid$^{2}$, H. V{\"o}lzke$^{2}$, B. Preim$^{1}$}
\authorfooter{
%% insert punctuation at end of each item
\item
 $^1$Otto-von-Guericke University Magdeburg, Germany.
\item
 $^2$Ernst-Moritz-Arndt-University Greifswald, Germany.
}

@@

Morphometric features are derived to allow for statistical comparison of the tissue, which incorporates mostly positions, diameters, volumes and relative distances and alignment to other structures.
% 
% - Data Format\\
% - Missing Data\\
% - Registration of Image Data (VMV'12)\\
% - Extraction of Image Parameter
\subsection{Analysis Workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/InteractionLoop}
 \caption{Detailed \emph{IVA Loop} as extension from Figure~\ref{fig:WorkflowComparison}.
 %
 Usually starting with a selection of a feature of interest (user-driven or via data mining techniques), the data are mapped using a visualization technique appropriate for the selected data types.
 %
 The data are visualized in the range and domain space, which can then be brushed, yielding new groups, to be investigated using further features.
 %
 Note that adjacent steps are directly connected via feedback loops allowing for an iterative refinement and giving as much freedom to the user as possible.}
 \label{fig:InteractionLoop}
\end{figure}
%-------------------------------- Figure --------------------------------
\begin{figure*}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 % \includegraphics[width=3.5in]{figures/visualization}
 %\includegraphics[width=1\textwidth]{figures/visualization}
 \includegraphics[width=1\textwidth, resolution=300]{figures/visualization}
 \caption{(Left) Screenshot from the front-end, which is divided as follows: (a) The sidebar containing all features as well as the groups defined in the analysis process; (b) the canvas area where features can be added via drag and drop and the visualization can be chosen automatically according to the data type; (c) the interactive pivot table showing the exact numbers for each displayed feature combination. The data displayed is used to analyze the lumbar spine. Features can be added freely on the canvas via drag and drop. Dropping the \emph{gender} parameter on the already plotted \emph{body size} container creates a mosaic plot combining both features (right).
 }
 \label{fig:visualization}
\end{figure*}
%-------------------------------- /Figure -------------------------------
Our proposed \emph{IVA} workflow knows three major steps as illustrated in Figure~\ref{fig:InteractionLoop}: Feature selection, visualization and brushing.
%
A hypothesis-driven analysis usually starts with the selection of features.
%
A feature selection from a shape-based clustering, which creates shape groups.
%
\emph{Hypothesis generation} with focus on image data starts with a shape-based clustering.
%
\com{Multivariate analysis using the adjacency matrix may have also be explained here, depending on the Application chapter.}
%
The feature is mapped using an automatically chosen visualization appropriate for its data type (described in detail in the following section).
%
The visualization techniques have to combine both image-and non-image data in order to set domain and range data in relation to each other.
%
In our system, the visualization can either be brushed or new features can be added to the analysis. 
%
Brushing methods are subdivided using the previously described \emph{IVA} patterns.
%
Brushed regions are treated like features as they divide the subject space just like categorical features.
%
Selecting features also triggers a \emph{multivariate analysis} using contingency values (described in the following section) to highlight features associated with selected features.
%
A sample workflow using interaction and visualization techniques described in the next section can be seen in Figure~\ref{fig:visualization}.

@@

The front-end is created using \texttt{Bootstrap}\footnote{Developed by Twitter, \url{http://getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%
Information visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} library \cite{D3}, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
%
The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by Nicolas Kruchten, \url{http://nicolas.kruchten.com/pivottable}}.
%
\texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} allows GPU-accelerated data rendering using \texttt{WebGL}.
%
Communication between client and server is enabled via the WebSockets protocol.
%
Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{http://mathworks.com}}, we had to access them using the \texttt{NodeJS} server.
%
We accomplished this by converting it to a parameterized standalone console application that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard out and returns it in a proper format to the client.
%
All parameter-steered console applications can be incorporated in this context.
\\\\
\section{Application}
This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
%
We conducted a case study with two domain experts on a data set compiled to analyze lower back pain. 
%
This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
%
Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
%
In comparable studies, only a few shape-related features are included \cite{Lang2011}.
%
To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
%
Determining risk factors in this area can lead to \cite{Fletcher2012}:
\begin{itemize}
	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
	\item prognostic features for diagnosis and treatment of lumbar back pain, and
	\item determination of particularly affected risk groups.
\end{itemize}
%
Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
%
\subsection{The Lumbar Spine Data Set}
We divide the data set into image and non-image data.
%
There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
%
The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
%
%The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.

@@

We conducted a qualitative case study with two epidemiological domain experts.
%
Both of them are practicing clinicians with focus on epidemiological research.
%
The first clinician is an specialist in internal medicine and also the leader of the SHIP.
%
One clinician is an radiologist and responsible for the image acquisition of the SHIP and has a deep understanding on the available image data.
%
\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule into account, the case study was done completely web based.
%
Since one major reason for the focus on web technologies, no software needed to be exchanged.
%
A website link was send to the users, which they could open with their browser of choice and start using the tool.
%
To make the user input comprehensible, we observed their monitor via screen sharing.
%
Communication was enabled via webcam supported voice over ip.
%
The total setup time including installing the screen sharing application was about five minutes.
%
Video-recordings of the sessions made it possible to evaluate the sessions later on.
\com{TODO: Name der Screensharing Software mit angeben (Teamviewer)?}
\paragraph{Procedure.}
At first, controlled mouse and keyboard of the clinicians pc and demonstrated the basic functionalities of the system.
%
This included the Heatmap, the correlation view, how to introduce parameter visualization, how the color coding of differences works and the pivot tables.
%
As they understood the concepts, we handed over the mouse and keyboard control and only observed from this point on.
%
The epidemiologists were given two tasks: one hypothesis-free analysis of the data and one starting with an assumption.

% \subsection{Exploratory Analysis of the Lumbar Spine Data Set}
\subsection{Case 1: Hypothesis-free analysis}
Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
%
With our tool, there are two ways to achieve this.
%
Performing a \emph{local investigation} by clustering all subjects or use the contingency matrix for a \emph{multivariate analysis}.

\paragraph{Analysis 1.}
%
The latter was chosen first by both domain experts.
%
%They pointed out the high potential they see in this visualization, because it allows to look at the data in an unbiased way.
They pointed out the benefits of the possibility to look at all variables in the context of each other.
%
To cite the radiologist, it "illumunates the black box of the data", making it possible to look at the data in an unbiased way.
%
Due to the dense mapping of information in the contingency matrix it was suggested to make this visualization full screen.
%
The radiologist was looking for correlations with shape related features in the data, finding that they correlate strong with each other as well as with leg pain, age, body height and hormone replacement therapy.
%
The internist inspected features which are not derived from the images, such as parameters strongly related with age, such as income, blood fat values or number of born children.
%
This also shows, that even though the visualization may treat every feature equally, the domain experts is more likely interested in features they are familiar with and have personal clinical experience with.
%
%Many correlations observed in this view

After this initial overview, the expert performed a \emph{multivariate analysis} by introducing features, such as age, waist circumference, weight or lumbar spine centerline curvature as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
%
+ Cramer test 
\com{Allgemein: Workflow noch einmal grob beschreiben - Overview first, details on demand}
%
\paragraph{Analysis 2.}
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/hypothesisfree}
 \caption{(a) Clustering result of all subjects. The bar chart height indicates the number of subjects in the cluster. The difference to the mean shape is color-coded, whereas blue represents no difference and red a large difference. (b) A pivot table relates cluster results to gender and employment status. Cluster 8 is striking because of the high proportion of women, while cluster 2, 3, 4 contain almost only men.}
 \label{fig:hypothesisfree}
\end{figure}
\com{ToDo: Rewrite}
%
The internist started also with a \emph{multivariate analysis} using the contingency table, but found no associations he was interested in.
%
Therefore, he applied the \emph{local investigation} pattern by a shape grouping step using shape-based clustering via dragging the "All subjects" feature into the canvas area, triggering the shape clustering.
%Shape-based hypothesis-free exploration starts with the \emph{local investigation} pattern by a shape grouping step using shape-based clustering.
%
The results for this step can be seen in Figure~\ref{fig:hypothesisfree} (a).

Cluster 9 represents subjects with average shape.
%
Other shapes differ with respect to size, such as cluster 2, 3, 8, 10 where the last one also represents a more straight spine, which is usual for subjects with larger body size.
%
Cluster 4, 5, 7 contain outliers, characterized by their unusual shape and small number.
%
Cluster 8 has the second largest number of elements and was therefore of special interest.
%
The user opened the contingency pane to performing a \emph{multivariate analysis} by looking at \emph{Cram\'{e}r's V} contingency values of cluster~8.
%
This revealed associations of this group with employment status, gender, body size, age, thyroid nodules and blood fat value.
%
Another \emph{multivariate analysis} using a pivot table set gender and employment status in relation to clustering affiliation\add{.} 
%
\add{It s}hows that cluster~8 contains mostly women and also has a larger unemployment rate (see Figure~\ref{fig:hypothesisfree}), while the overall employment rate of women and men in the data set is almost exactly 50\%.
%
While all observed features seem to be plausible associations related to back pain, the values indicate that cluster~8 contains subjects with chronic back pain radiating to the legs.
%
Metabolic parameters, such as blood fat and blood sugar, are also possibly associated features.
%
The employment status is a feature relating to many different lifestyle factors such as income or nutrition as well as age and might act as confounder.

@@

\subsection{Case 2: Hypothesis-based analysis}
If the user proposes a hypothesis about a relation between a non-image feature regarding shape, the workflow slightly differs from the hypothesis-free analysis.
%
The starting point follows the \emph{feature investigation} pattern, where a feature of interest is selected by dragging it into the canvas area and viewing the subject's distribution as well as their shape differences.
\paragraph{Analysis 1}
The radiologist performed a \emph{feature investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
%
Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
%
The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
%

\paragraph{Analysis 2.}
Hypothesis: "Back pain is associated with age and therefore there differences in the lumbar spine shape are observable." \com{TODO Image Mosaic Plot Age, Backpain}.
To validate this hypothesis, a \emph{multivariate analysis} was performed by introducing the dichotomous feature "\emph{Did you experience backpain in the last three moth?}" together with the age as mosaic plot by dropping both features on the canvas area.
%
The experts were not able to observe the expected effect in the visualization.
%
Reasons for this are twofold.
%
For one, age influences the lumbar spine shape very heavily, as seen in Fig~\ref{?}, while the differences between subjects with and without back pain are small.
%
The major difference seen in the visualization are therefore presented by the age feature, masking differences via the back pain parameter.

The second explanation is the commonality of back pain in our society.
%
As seen in Fig~\ref{?}, subjects reporting back pain are even the majority, which makes it difficult to extract parameters which reliably describe back pain.
%
A \emph{local investigation} using the contingency table showed a strong association between back pain with gender and body height.
%
Body height was explained to be a confounder for gender, since female subjects smaller on average than male subjects.
%
The analysis solely based on shape only accentuated gender differences, not in back pain.
%This makes it hard to analyze shape related features for this disease
%
A next step would be to use a variable describing the strength of back pain and only analyzing strong back pain!
%

\com{ToDo: Ability to better create sub groups would be useful as well as better segmentation}
%gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview

%
%\com{Video 00.29. Analysis of Gender and Age - older subjects are closer to females, younger to males. This is due to shrinking of the spine in the aging process.}
%\com{Backpain three month > age. There are more subjects reporting backpain than those who dont. Since influences heavily the average shape - since most people have back pain, the average shape already is associated with back pain, which makes it difficult to extract parameters which reliably describe back pain. (1.05) The radiologist pointed out that for the mosaic view it would be beneficial to abstract the shape and curvature and categorize it into shapes s, i form. It was also pointed out that in the plot back pain>age, back pain can not be differenciated because the difference is mainly encoded via the age, more variables need to be added (gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview).}

\paragraph{Analysis 3}
\com{ToDo: Needs to be rewritten}
\begin{figure}[htb]
 \centering
 %\includegraphics[width=1.5in]{figures/InteractionLoop}
 \includegraphics[width=3.5in]{figures/hypothesisbased}
 \caption{(a) Dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
 %
 Mean shapes between the groups show no difference.
 %
 (b) Shape-based clustering for all subjects who suffered from back pain yields 6 groups. 
 %
 Note that the difference in subject count is due to the missing shape information for some subjects.
 %
% The total count of subjects in (b) is smaller 
 }
 \label{fig:hypopthesisbased}
\end{figure}
%
In our use case, epidemiologists were interested in the dichotomous questionnaire answer to ``\emph{Did you experience back pain in the past three months?}''.
%
The mean shapes of the resulting visualization show no difference between the two groups (see Figure~\ref{fig:hypopthesisbased} (a)).
%
Either there are no differences or the variance information was lost in the mean shape calculation.
%
Since the focus is on subjects suffering from back pain, the clustering results of these subjects are then drawn into the canvas area, yielding six clusters as seen in Figure~\ref{fig:hypopthesisbased} (b).
%
Cluster 5 stood out for having a so-called \emph{hyperlordosis}, a strong curvature of the lumbar spine associated with back pain.
